---
title: "SA with Python and sequenzo"
description: |
  Running sequenzo from R using reticulate
output: distill::distill_article
---


```{r setup, include=FALSE}
# Load required packages
library(here)
pacman::p_load(formatR ,icons, kableExtra, reticulate, 
               scales, tictoc, tidyverse, TraMineR)

# Output options
knitr::opts_chunk$set(
  eval=TRUE, 
  echo=TRUE)


options("kableExtra.html.bsTable" = T)

# load family biographies
family <- readRDS(here("data", "familybio.rds"))

# # Only if necessary: Install Python and required libraries
# install_python()
# py_install("sequenzo")

# Import Python modules
sequenzo <- import("sequenzo")

```

<style>
pre {
  white-space: pre-wrap !important;
  word-wrap: break-word !important;
  overflow-wrap: break-word !important;
  max-width: 100%;
  overflow-x: auto;
}

.sourceCode {
  white-space: pre-wrap !important;
  word-wrap: break-word !important;
}

pre code {
  white-space: pre-wrap !important;
  word-wrap: break-word !important;
}
</style>


```{r, xaringanExtra-clipboard, echo=FALSE}
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clone fa-2x\" style=\"color: #301e64\"></i>",
    success_text = "<i class=\"fa fa-check fa-2x\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times fa-2x\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

After our book was published, [Yuqi Liang](https://www.yuqi-liang.tech/){target="_blank"} and her team developed [`Sequenzo`](https://sequenzo.yuqi-liang.tech/en/){target="_blank"}, a `r fontawesome("python")` package for social sequence analysis. Beyond making sequence analysis more accessible to Python users, this package offers significant advantages due to Python's powerful computing capabilities: it is considerably faster than available R tools and better suited for larger datasets. Since we are not Python users ourselves, this brief introduction is specifically aimed at R users who want to leverage the power of `Sequenzo` from within R (as part of an R script) using the `{reticulate}` package.


## Setup   

If you have not installed Python and sequenzo yet you have to run the following two `{reticulate}` functions first:


```{r, echo=TRUE, eval=FALSE}
# use (and install if necessary) pacman package 
if (!require("pacman")) install.packages("pacman")
library(pacman)

# load and install (if necessary) required packages for this course
pacman::p_load(
  kable,      # Table
  reticulate, # R interface to Python
  tictoc,     # for measuring the duration of distance computation
  tidyverse,  # universal toolkit for data wrangling and plotting 
  TraMineR    # The sequence analysis toolkit for R
  )

# Install Python and sequenzo using reticulate
install_python()
py_install("sequenzo")
```

Now you are ready to import the data for example application and `sequenzo`.

```{r, echo=TRUE, eval=FALSE}
# Import Python modules
sequenzo <- import("sequenzo")


# load example data: family biographies from PAIRFAM
family <- readRDS("familybio.rds")
```

## Preparing the data

The imported data frame `family` contains sequence data from `r label_comma()(nrow(family))` respondents of the German Family Panel (pairfam). The 264 sequence variables are numbered and start with the prefix `state`. They provide monthly information on family biographies - a combination of partnership status and parity—from age 18 to 40.

```{r echo=FALSE}
shortlab.family <- c("S", "Sc", 
                     "LAT", "LATc", 
                     "COH", "COHc",
                     "MAR", "MARc1", "MARc2+")

longlab.family <- 
  c("Single, no child", "Single, child(ren)",
    "LAT, no child", "LAT, child(ren)", 
    "Cohabiting, no child", "Cohabiting, child(ren)", 
    "Married, no child", "Married, 1 child", "Married, 2+ children")

kable(tibble(num = 1:9,
             state = longlab.family, 
             shortlabel = shortlab.family), 
      col.names = c("#", "State", "Short Label")) %>%
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed"),
                full_width = F)
```



### Step 1: Define a sequence object in R

```{r echo=TRUE}
# define long and short labels for sequence vars
shortlab.family <- c("S", "Sc", 
                     "LAT", "LATc", 
                     "COH", "COHc",
                     "MAR", "MARc1", "MARc2+")

longlab.family <- 
  c("Single, no child", "Single, child(ren)",
    "LAT, no child", "LAT, child(ren)", 
    "Cohabiting, no child", "Cohabiting, child(ren)", 
    "Married, no child", "Married, 1 child", "Married, 2+ children")

# define sequence object in TraMineR
family.seq <- seqdef(data = select(family, starts_with("state")),
                     states = shortlab.family,
                     labels = longlab.family, 
                     alphabet = 1:9,
                     id = family$id)

```

### Step 2: Define a sequence object with `sequenzo` using `{reticulate}`

```{r echo=TRUE}
# get data into recommended format
seqdata <- family |> 
  mutate(across(everything(), as.character)) |> 
  rename_with(~ str_remove_all(.x, "state"))
  
# sequence data to pyhton 
df_py <- r_to_py(seqdata)

# set parameters
time_list <- as.character(1:264)
states <- as.character(1:9)
labels <- longlab.family

# define sequence data in sequenzo
dataset <- sequenzo$SequenceData(
  df_py,
  time = time_list,
  time_type = "age",
  id_col = "id",
  states = states,
  labels = labels
  )
```

## Compute dissimilarities & compare performance

Now we are set to compute the dissimilarity matrices. For this example we choose Optimal Matching with indel costs of 1 and constant substitution costs of 2. 

```{r eval=TRUE, echo=FALSE, message=FALSE, results='hide', cache=TRUE}
# TraMineR ----

tic()

# OM with TraMineR
om.const<-seqdist(family.seq,
                  method = "OM", 
                  indel = 1,
                  sm = "CONSTANT",
                  norm = "none")

time_traminer_sec <- toc(quiet = TRUE)
time_traminer_sec <- time_traminer_sec$toc - time_traminer_sec$tic


# sequenzo ----

tic()

# OM with sequenzo
om.sequenzo <- sequenzo$get_distance_matrix(
  seqdata = dataset, 
  method = "OM",
  sm = "CONSTANT",
  norm = "none",
  full_matrix = TRUE
)

time_sequenzo_sec <- toc(quiet = TRUE)
time_sequenzo_sec <- time_sequenzo_sec$toc - time_sequenzo_sec$tic


min_traminer <- round(time_traminer_sec/60,2)
min_sequenzo <- round(time_sequenzo_sec/60,2)

```



```{r eval=FALSE}
# OM with TraMineR
om.const<-seqdist(family.seq,
                  method = "OM", 
                  indel = 1,
                  sm = "CONSTANT",
                  norm = "none")

# OM with sequenzo
om.sequenzo <- sequenzo$get_distance_matrix(
  seqdata = dataset, 
  method = "OM",
  sm = "CONSTANT",
  norm = "none",
  full_matrix = TRUE
)
```

As indicated by the developers, `sequenzo`'s `get_distance_matrix` is notably faster `r fontawesome("rocket")` than `TraMineR`'s `seqdist`. On our test machine (Surface Pro 7+, 11th Gen Intel Core i7-1165G7 @ 2.80GHz, 4 cores; 16 GB RAM), `TraMineR` requires **`r min_traminer` minutes**, whereas `sequenzo` needs only **`r min_sequenzo` minutes** (`r label_percent()(min_sequenzo/min_traminer)` of the time) to compute the distances.

## Check results

Finally, we confirm that the results from both packages are identical.

```{r}
# Visual inspection
om.sequenzo[1:5,1:5]
om.const[1:5,1:5]

# Test if distances are the same
all.equal(as.matrix(om.sequenzo), as.matrix(om.const))

```
