---
title: "Measuring Associations Between Sequences and Covariates II"
description: |
  Chapter 6.2.2 Bayesian Information Criterion and the Likelihood Ratio Test
output: distill::distill_article
bibliography: Companion.bib
csl: https://www.zotero.org/styles/apa
---

```{r setup, include=FALSE}
# Load required packages
library(here)
source(here("source", "load_libraries.R"))

# Output options
knitr::opts_chunk$set(eval=TRUE, echo=TRUE)
options("kableExtra.html.bsTable" = T)

# load data for Chapter 2
load(here("data", "6-0_ChapterSetup.RData"))

# source code required to produce table shown in the book
source(here("source", "6-2_Table_6-4_Discrepancies-BIC.R"))

```

```{r, xaringanExtra-clipboard, echo=FALSE}
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clone fa-2x\" style=\"color: #301e64\"></i>",
    success_text = "<i class=\"fa fa-check fa-2x\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times fa-2x\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

<details><summary>**Click here to get instructions...**</summary>

- Please download and unzip the replication files for Chapter 6
([`r fontawesome::fa("far fa-file-zipper")` Chapter06.zip](source/Chapter06.zip)). 
- Read `readme.html` and run `6-0_ChapterSetup.R`. This will create `6-0_ChapterSetup.RData` in the sub folder `data/R`. This file contains the data required to produce the table shown at the bottom of this page. 
- We also recommend to load the libraries listed in the Chapter 6's `LoadInstallPackages.R`

```{r, eval=FALSE}
# assuming you are working within .Rproj environment
library(here)

# install (if necessary) and load other required packages
source(here("source", "LoadInstallPackages.R"))

# load environment generated in "6-0_ChapterSetup.R"
load(here("data", "R", "6-0_ChapterSetup.RData"))

```
</details>

\


**Table 6.4** in Chapter 6.2.2 presents group comparisons based on a discrepancy analysis and an alternative approach recently proposed by Liao and Fasang [-@LiaoFasang2021]. In this example we examine the labor market trajectories stored in `activity.year.seq`. The group comparisons are using dummy variables stored in `activity`: `east` (living in East vs West Germany), `sex` (male vs. female),  `highschool` (at least highschool degree: yes vs. no). 

The analysis are done with [`{TraMineR}`](http://traminer.unige.ch){target="_blank"}'s `dissassoc` function and the `seqCompare` function from the [`{TraMineRextras}`](http://traminer.unige.ch){target="_blank"} package.

In the book we present the following table (note that we highlighted one line we are going to elborate on in the following paragraphs):


```{r, echo=FALSE, eval=TRUE}
bic.pseudoR2 %>%
  kbl(digits = c(2,3,2,3), align = c("r", rep("c", 3), "r")) %>%
  row_spec(3, background = "#8DCEA4") %>% 
  row_spec(0, italic = T) %>%
  kable_styling(full_width = F)  %>%
  pack_rows("West vs East", 1, 3, indent = F) %>%
  pack_rows("Men vs Women", 4, 12, indent = F) %>%
  add_indent(c(2,3,5,6,8, 9, 11, 12)) %>%
  add_header_above(c(" " = 2, 
                     "Likelihood Ratio Test" = 1, 
                     "Discrepancy Analysis" = 2, " " = 1))
```

Obviously, the table includes a lot of group comparisons and it would require a lot of repetitive coding if each single comparisons would be explicitly spelled out. For that reason, we were using [`{purrr}`](https://purrr.tidyverse.org/){target="_blank"}'s `map` function to to do the comparisons within loops. We also wrote several lines of code for extracting and formatting the results to be displayed in a nice table  using `{knitr}`'s `kable` and the `{kableExtra}` package. In the accompanying script file `6-2_Table_6-4_Discrepancies-BIC.R` stored in [`r fontawesome::fa("far fa-file-zipper")` Chapter06.zip](source/Chapter06.zip)) you can find the full code required to produce the table shown above.

On this page, however, we do not further elaborate on all these technicalities and rather focus on briefly showcasing the plain [`{TraMineR}`](http://traminer.unige.ch){target="_blank"} commands required to obtain the results shown in the table. As you see, this does not require a lot of coding:  

We illustrate the procedure by calling the functions required to obtain the figures shown in the highlighted row of table displayed above. The functions `dissassoc` and `seqCompare`are easy to use. The most "tricky" part for R novices is probably to apply the functions to the correct subset of the data. In our example we are interested in the question of whether the sequences of East German women are systematically different from the sequences of their West German counterparts. Hence, our analysis have to be restricted to women. While our sequences are stored in the object `activity.year.seq` any further information characterizing the individual who experience these sequences is stored in the dataframe `activity`. We start our analysis by inspecting how many women reside in East and West Germany. (`table(activity$east[activity$sex==1])`).   

```{r, echo=FALSE}
activity %>%
  filter(sex==1) %>% 
  count(east) %>% 
  as.data.frame() %>% 
  set_rownames(c("West Germany", "East Germany")) %>% 
  select(Women = n) %>% 
  kable() %>% 
  kable_styling(full_width = FALSE)
```

We then proceed by generating a logical vector (`women`) indicating which of the `r nrow(activity)` rows of `activity` is referring to a female respondent.    


```{r echo=TRUE, eval=TRUE}
women <- activity$sex==1

# visual inspection
activity %>% 
  select(sex) %>% 
  mutate (women = women)

```

We use this vector to subset women from the dissimilarity matrix `activity.year.om`, the sequence object `activity.year.seq` and the vector `activity$east` when we call `dissassoc` and `seqCompare`. Note that subsetting the pairwise distance matrix requires to extract the correct rows **and** columns `[women,women]`, while subsetting `activity.year.seq` only requires to extract the desired rows `[women, ]`. 

Below we compute the values shown in the table. While the discrepancy analysis requires a dissimilarity matrix (`activity.year.om[women,women]`) as an input, `seqCompare` computes the dissimilarity matrix on its own and requires a sequence object (`activity.year.seq[women,]`) and instructions on how to compute the distances (`method="OM", sm="CONSTANT"`) as an input. If you want to compare the results of the two approaches, the distances should be computed with the same method. Both functions require a vector (`activity$east[women]`)specifying the groups to be compared. Note that `seqCompare` currently only accepts a dichotomous grouping indicator.  


```{r}
discr.region <- dissassoc(activity.year.om[women,women], 
                          group = activity$east[women])

bic.region <- seqCompare(activity.year.seq[women,], 
                         group = activity$east[women],
                         method="OM", sm="CONSTANT")

```

The output of `seqCompare` is quite parsimonious. For **Table 6.4** we extracted the p-value (`bic.region[,2]`) and the BIC difference (`bic.region[,3]`)

```{r}
bic.region %>%
  kable(digits = c(2,3,2,2,2)) %>% 
  column_spec(2:3, background = "#8DCEA4") %>% 
  kable_styling(full_width =FALSE) 
```

`dissassoc`, on the other hand, produces quite a lot of output which can be stored as a list object. The results shown in the table are the total number of cases (`discr.region$groups[3,1]`), the pseudo-$R^2$ (`discr.region$stat[3,1]`), and the corresponding p-value (`discr.region$stat[3,2]`). 

We close this section double checking if the subsetting worked as intended by inspecting the case numbers stored in `discr.region$groups`. The first row refers to East German women and the second one to West German women. The case numbers are as expected (see frequency table above). Hence, we can be confident that we achieved what we wanted. 

```{r}
kable(discr.region$groups) %>% 
    kable_styling(full_width =FALSE) 

```

For the more complex subsetting including all groups shown in **Table 6.4** we refer you to the script `6-2_Table_6-4_Discrepancies-BIC.R`. There we also illustrate how to extract, arrange, and format the outputs from `dissassoc` and `seqCompare` to produce the table shown at the beginning of this page.
