---
title: "Crisp clustering algorithms and cluster quality indeces"
description: |
  Chapter 4.1	Clustering sequences to uncover typologies
output: distill::distill_article
---

```{r setup, include=FALSE}

# Load required packages
library(here)
source(here("source", "load_libraries.R"))

# Output options
knitr::opts_chunk$set(eval=TRUE, echo=TRUE)
options("kableExtra.html.bsTable" = T)

# load data for Chapter 4
load(here("data", "4-0_ChapterSetup.RData"))

```


```{r, xaringanExtra-clipboard, echo=FALSE}
htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clone fa-2x\" style=\"color: #301e64\"></i>",
    success_text = "<i class=\"fa fa-check fa-2x\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times fa-2x\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

<details><summary>**Click here to get instructions...**</summary>

- Please download and unzip the replication files for Chapter 4
([`r fontawesome::fa("far fa-file-archive")` Chapter04.zip](source/Chapter03.zip)). 
- Read `readme.html` and run `4-0_ChapterSetup.R`. This will create `4-0_ChapterSetup.RData` in the sub folder `data/R`. This file contains the data required to produce the plots shown below. 
- You also have to add the function `legend_large_box` to your environment in order to render the tweaked version of the legend described below. You find this file in the `source` folder of the unzipped Chapter 4 archive.
- We also recommend to load the libraries listed in Chapter 4's `LoadInstallPackages.R`

```{r, eval=FALSE}
# assuming you are working within .Rproj environment
library(here)

# install (if necessary) and load other required packages
source(here("source", "load_libraries.R"))

# load environment generated in "3-0_ChapterSetup.R"
load(here("data", "R", "3-0_ChapterSetup.RData"))

```
</details>

\

In chapter 4.1, we introduce....

## 

```{r, eval=TRUE, echo=TRUE}
```
